%!TEX root = predictability.tex

\begin{abstract}
Transaction processing is an important part of modern database management
systems(DBMSs), so a lot of work has been done to improve the performance of
it. However, few people pay attention to the predictability of performance,
which is fundamental to many application of database systems such as
in-cloud database services that provide performance guarantee. The service
providers will not be able to provide such kind of guarantees unless their
services are based on database systems with predictable performance.

In this paper, we show that variance of transaction latency is a very
severe problem in traditional database systems like MySQL. We propose a
profiling framework called VProfiler that uses the source code of a database
system to identify the main sources of variance in transaction latency. This
framework works by breaking down the variance of latency into variances and
covariances of the execution time of the functions in the source code, which is
the best source of information about the actual cause of variance. Using MySQL
as an example, we show its main sources of variance found using our profiling
framework, and propose two methods to reduce the variance of latency after
close investigation of these sources. Experiment results show that these
two methods, when combined, achieve a 42.3\% reduction in 99th percentile.
Besides these methods, we also demonstrate how using faster storage devices and
tuning the configuration of MySQL can further improve its performance
predictability.
\end{abstract}

